Supporting information: Copulas and their potential for ecology

### 2019-09-07 Paper.Rmd
some progress writing a Background section

<!--DONE ***DANr: reword this para so people don't think the ranks *are* the 
copulas, and to try to pre-defend against some of what ref 1 said. Also,
state that these are conceptual introductions only and are not formal,
a more comprehensive introduction is in xyz-->
Well-developed approaches do exist, however, and are 
applied widely in other fields, 
that provide a description of the dependence 
between two or more variables that can be mathematically proven 
to be complete [@joe2014_dependence; @nelsen2006_copula]. 
The approaches are based on 
the concept of the *copula*; copula approaches separate the information of a bivariate 
random variable into two distinct parts: the information in the 
marginal distributions (which says nothing about the dependence between the 
variables), and the remaining information (which is solely about
dependence). We here introduce some simple concepts, based on taking ranks,
that relate to copulas and give a conceptual 
flavor of copula ideas to be introduced 
more formally in the section on Background below. Given 
a sample $(x_i,y_i)$, $i=1,2,\ldots,n$ (e.g., 
Fig. \ref{fig_cop_pedag1}A-B), information about the structure of the dependence
between $x$ and $y$ can be separated from information contained 
in the marginal distributions by 
considering the bivariate plot of $u_i$ versus $v_i$, where
$u_i$ is the rank of $x_i$ in the set $x_j$ ($j=1,\ldots,n$), divided by
$n+1$; and $v_i$ is defined in the same way but using the $y_i$. The rank
of the smallest element in a set is $1$. The $u_i$ and 
$v_i$ are called *normalized ranks* of the $x_i$ and $y_i$; they relate to 
the empirical cummulative distribution function (cdf) of the $x_i$ and $y_i$,
respectively. See, for instance,
Fig. \ref{fig_cop_pedag1}C-D (the points), which show the normalized ranks of 
A-B, respectively. Ranking makes the marginal distributions of 
the component datasets uniform, isolating the 
dependence structure. Dependence structure and marginals can then be studied 
separately. Normalized rank plots such as Fig. \ref{fig_cop_pedag1}C-D}
relate to copulas in a way to be described in the section on Background below, 
after a definition of copulas
is provided. Note that Pearson correlation, even though it is the most commonly
used measure of relationships between variables, is modified by monotonic
tranformation of the component variables (Appendix 
\ref{SM-pedagog1_details}, Fig. \ref{SM-fig_transf_pearson_spearman}),
and therefore reflects not only dependence information, but also
information on the marginals. Spearman and Kendall correlations, being rank-based, 
are not influenced by monotonic transformations of the component variables; 
they provide information solely on the dependence between those variables.

@@

The goal of this paper is to explore the potential for 
applications of copulas in ecology. We provide a sketch of an introduction
to copula concepts in the Background section below, but this is a sketch
only - readers can find comprehensive introductions elsewhere
[@nelsen2006_copula; @joe2014_dependence]. Instead our goal is to 
help determine to what extent ecological understanding may benefit from
using copulas to help understand complex dependence 
structures between variables.
Our results provide support for the hypothesis that environmental, 
ecological and evolutionary processes 
generate and transmit complex dependence structures, including tail dependence, 
with consequences for theory and 
applications. We hope our results will stimulate further
evaluation of this hypothesis by other researchers.

@@

<!--DONE ***DANr: Maybe use the words tail dependence more and the word copula
less in the below few paras? Generally the words "copula structure" may be rubbing some
people wrong. Replacement words? What do you actually mean here? Maybe call it
asymmetric dependence, or complex dependence structure?-->
One way we may expect, *a priori*, a copula 
study of complex dependence structures between
ecological variables
to better reflect ecological relationships relates to Liebig's law of the
minimum (the idea that growth is controlled not by total resources but by the scarcest 
resources). If, for instance, the growth of a plant depends on soil 
nitrogen, N, and other factors, a plot of plant growth rates versus soil N may 
look more like Fig. \ref{fig_cop_pedag1}A or C than like 
Fig. \ref{fig_cop_pedag1}B or D. Nitrogen 
will control plant growth, producing a clear
relationship, only when it is limiting, in the left portions of the distributions. 
The portions of the distributions that dominate the association is visible in rank-based
plot, but may not be revealed by standard correlation techniques.

@@

<!--DONE ***DANr: Need some caveating here to accommodate the people who slavishly
follow the old maxim that correlation is not causation.-->
A third reason we may expect an understanding of complex 
dependence structures to benefit 
ecological research is that such an approach may help illuminate 
the causal mechanisms between variables. For instance, if two species
Sp1 and Sp2 are strong competitors, abundances of the two species
across quadrats should be negatively related, as in 
Fig. \ref{fig_cop_pedag1}E-F. If Sp1 is plotted on the horizontal axis
and Sp2 on the vertical, then Fig. \ref{fig_cop_pedag1}E 
may suggest Sp1 is the dominant 
competitor: when Sp1 is abundant, Sp2 is necessarily rare because it is suppressed; 
whereas when Sp1 is rare, Sp2
may be abundant, or may also be relatively rare due to 
limiting factors other than Sp1. 
Alternatively, Fig. \ref{fig_cop_pedag1}F may suggest that Sp2
is the dominant competitor. Although other causal hypotheses may produce 
similar dependence structure, and it is usually impossible to obtain
complete information on causal pathways from analyses which are
fundamentally based on association data, more detailed 
examination of the patterns of dependence of variables may
well be combined with biological and other reasoning to rule out
some causal hypotheses which could not be eliminated solely 
with standard correlation approaches.

@@

We approached our overall goal of exploring the potential of copula approaches 
for ecology by addressing the following specific questions. (Q1) Do datasets
in ecology have dependence structure distinct from that of a multivariate 
Gaussian/normal
distribution (here called non-normal copula structure)? Do they show tail 
dependence distinct from that of a normal distribution, and in particular
do they show asymmetric tail dependence? Normal copula structure is assumed
by standard approaches that use multivariate 
normal distributions or distributions obtained by transforming
the marginals of a normal distribution. So Q1 asks whether ecological
data contain dependency information distinct from what can be modelled with
standard approaches. (Q2) If the answer to
Q1 is "yes," what are some possible causes/mechanisms of non-normal
copula structure and tail dependence? And (Q3) what are the 
consequences of non-normal copula
structure and tail dependence for ecological understanding and
applications? After a Background section (the second section), we address Q1 
(third through fifth sections of the paper) 
by analyzing several datasets, including 
environmental, species-trait, phenological, population, community, and 
ecosystem functioning data, selected to span multiple 
sub-fields and organizational levels of ecology. 
We address Q2 (sixth and seventh sections of the paper), using 
simple models. We address Q3 (eighth and ninth sections of the paper) using 
both data and models. The final section is the Discussion.
Multiple analyses are brought to bear on each question, and these
are summarized diagramatically in Fig. \ref{MasterFigure}, which can 
serve as a *post hoc* guide to the paper. 
Copula approaches have been used to great effect in neuroscience [@onken2009], 
bioinformatics [@Kim2008], medical research [@Emura2016], direct study of environmental 
variables [@Serinaldi2008; @She2018; @Goswami2018; @li2013], and finance [@Li2000],
but have been used only rarely in ecology. 
We argue that benefits of using a complete 
copula description of dependence in ecology will be substantial.

@@

<!--***DANr: Add a new section above this one called Background on copulas.
It can have the following kinds of stuff (probably not exhaustive, and probably
not in order, and some of this material is currently below):
The definition
The explanation of how ranks relate to estimating the copula parameters
How to estimate carry out fitting (basically, MLE fitting on ranks)
What some of the different copula families look like
What tail dependence actually is
-->

<!--***DANr: Maybe move the below two sentences into the new section on Background-->
See @nelsen2006_copula and @joe2014_dependence
for general background. A theorem of @sklar1959_theorem
states that copulas give a unique and complete 
representation of the dependence
between variables.

@@

Conceptually, a bivariate copula is a random variable 
on the unit square, $[0,1] \times [0,1]$, with uniform maginals. 
If $(X,Y)$ is a continuous random
variable and the cummulative distribution functions of $X$ and $Y$ are 
$F_X(x)=P[X\leq x]$ and $F_Y(y)=P[Y\leq y]$, then the dependence structure
of $(X,Y)$ is revealed by the random variable $(F_X(X),F_Y(Y))$,
which takes values in $[0,1] \times [0,1]$ and has 
uniform marginals.

>>>>

<!--DONE ***DANr: reword this para so people don't think the ranks *are* the 
copulas, and to try to pre-defend against some of what ref 1 said. Also,
state that these are conceptual introductions only and are not formal,
a more comprehensive introduction is in xyz-->
Well-developed approaches do exist, however, and are 
applied widely in other fields, 
that provide a description of the dependence 
between two or more variables that can be mathematically proven 
to be complete [@nelsen2006_copula; @joe2014_dependence; @MaiScherer2017]. 
The approaches are based on 
the concept of the *copula*; copula approaches separate the information of a bivariate 
random variable into two distinct parts: the information in the 
marginal distributions (which says nothing about the dependence between the 
variables), and the remaining information (which is solely about
dependence). We here introduce some simple concepts, based on taking ranks,
that relate to copulas and give a conceptual 
flavor of copula ideas to be introduced 
more formally in the section on Background below. Given 
a sample $(x_i,y_i)$, $i=1,2,\ldots,n$ (e.g., 
Fig. \ref{fig_cop_pedag1}A-B), information about the structure of the dependence
between $x$ and $y$ can be separated from information contained 
in the marginal distributions by 
considering the bivariate plot of $u_i$ versus $v_i$, where
$u_i$ is the rank of $x_i$ in the set $x_j$ ($j=1,\ldots,n$), divided by
$n+1$; and $v_i$ is defined in the same way but using the $y_i$. The rank
of the smallest element in a set is $1$. The $u_i$ and 
$v_i$ are called *normalized ranks* of the $x_i$ and $y_i$; they relate to 
the empirical cummulative distribution function (cdf) of the $x_i$ and $y_i$,
respectively. See, for instance,
Fig. \ref{fig_cop_pedag1}C-D (the points), which show the normalized ranks of 
A-B, respectively. Ranking makes the marginal distributions of 
the component datasets uniform, isolating the 
dependence structure. Dependence structure and marginals can then be studied 
separately. Normalized rank plots such as Fig. \ref{fig_cop_pedag1}C-D
relate to copulas in a way to be described in the section on Background below, 
after a definition of copulas
is provided. Note that Pearson correlation, even though it is the most commonly
used measure of relationships between variables, is modified by monotonic
tranformation of the component variables (Appendix 
\ref{SM-pedagog1_details}, Fig. \ref{SM-fig_transf_pearson_spearman}),
and therefore reflects not only dependence information, but also
information on the marginals. Spearman and Kendall correlations, being rank-based, 
are not influenced by monotonic transformations of the component variables; 
they provide information solely on the dependence between those variables.

@@

The goal of this paper is to explore the potential for 
applications of copulas in ecology. We provide a basic introduction
to copula concepts in the Background section below, but this is a sketch
only - readers can find abbreviated [@Anderson2018] 
and comprehensive introductions elsewhere
[@nelsen2006_copula; @joe2014_dependence; @MaiScherer2017]. Instead our goal is to 
help determine to what extent ecological understanding may benefit generally from
using copulas. Copulas have already been used very effectively 
in a few studies to illuminate ecological phenomena 
[@Valpine2014; @Anderson2018; @Popovic2018], but such usage is
still rare. Our results suggest that environmental, 
ecological and evolutionary processes 
may commonly generate and transmit complex dependence structures, including tail 
dependence, and that greater use of copulas can illuminate the underlying processes.

@@

<!--DONE ***DANr: Maybe use the words tail dependence more and the word copula
less in the below few paras? Generally the words "copula structure" may be rubbing some
people wrong. Replacement words? What do you actually mean here? Maybe call it
asymmetric dependence, or complex dependence structure?-->
One way we may expect, *a priori*, a copula 
study of complex dependence structures between
ecological variables
to better reflect ecological relationships relates to Liebig's law of the
minimum. Liebig's law is 
the idea that growth is controlled not by total resources but by the resources which
are scarcest relative to organism needs. 
If, for instance, the growth of a plant depends on soil 
nitrogen, N, and other factors, a plot of plant growth rates versus soil N may 
look more like Fig. \ref{fig_cop_pedag1}A or C than like 
Fig. \ref{fig_cop_pedag1}B or D. Nitrogen 
will control plant growth, producing a clear
relationship, only when it is limiting, in the left portions of the distributions. 
The portions of the distributions that dominate the association are visible in a rank-based
plot, but may not be revealed by standard correlation techniques.

@@

<!--DONE ***DANr: Need some caveating here to accommodate the people who slavishly
follow the old maxim that correlation is not causation.-->
A third reason we may expect an understanding of complex 
dependence structures to benefit 
ecological research is that such an approach may help illuminate 
the causal mechanisms between variables. For instance, if two species
Sp1 and Sp2 are strong competitors, abundances of the two species
across quadrats should be negatively related, as in 
Fig. \ref{fig_cop_pedag1}E-F. If Sp1 is plotted on the horizontal axis
and Sp2 on the vertical, then Fig. \ref{fig_cop_pedag1}E 
may suggest Sp1 is the dominant 
competitor: when Sp1 is abundant, Sp2 is necessarily rare because it is suppressed; 
whereas when Sp1 is rare, Sp2
may be abundant, or may also be relatively rare due to 
limiting factors other than Sp1. 
Alternatively, Fig. \ref{fig_cop_pedag1}F may suggest that Sp2
is the dominant competitor. Although other causal hypotheses may produce 
similar dependence structure, and it is usually impossible to obtain
complete information on causal pathways from analyses which are
fundamentally based on associations, more detailed 
examination of the patterns of dependence of variables may
well be combined with biological and other reasoning to rule out
some causal hypotheses which could not be eliminated solely 
with standard correlation approaches. @Popovic2018 has previously
used copulas to help illuminate causal aspects of relationships
between species.

@@

We approached our overall goal of exploring the potential of copula approaches 
for ecology by addressing the following specific questions. (Q1) Do datasets
in ecology have dependence structure distinct from that of a multivariate 
Gaussian/normal
distribution (here called non-normal copula structure)? Do they show tail 
dependence distinct from that of a normal distribution, and in particular
do they show asymmetric tail dependence? Normal copula structure is assumed
by standard approaches that use multivariate 
normal distributions or distributions obtained by transforming
the marginals of a normal distribution. So Q1 asks whether ecological
data contain dependency information distinct from what can be modelled with
standard approaches. (Q2) If the answer to
Q1 is "yes," what are some possible causes/mechanisms of non-normal
copula structure and tail dependence? And (Q3) what are the 
consequences of non-normal copula
structure and tail dependence for ecological understanding and
applications? After a Background section (the second section), we address Q1 
(third through fifth sections of the paper) 
by analyzing several datasets, including 
environmental, species-trait, phenological, population, community, and 
ecosystem functioning data, selected to span multiple 
sub-fields and organizational levels of ecology. 
We address Q2 (sixth and seventh sections of the paper), using 
simple models. We address Q3 (eighth and ninth sections of the paper) using 
both data and models. The final section is the Discussion.
Multiple analyses are brought to bear on each question, and these
are summarized diagramatically in Fig. \ref{MasterFigure}, which can 
serve as a *post hoc* guide to the paper. 
Copula approaches have been used to great effect in neuroscience [@onken2009], 
bioinformatics [@Kim2008], medical research [@Emura2016], direct study of environmental 
variables [@Serinaldi2008; @She2018; @Goswami2018; @li2013], and finance [@Li2000],
but have been used only rarely so far in ecology (but see @Valpine2014, 
@Anderson2018 and @Popovic2018). 
We argue that benefits wider usage of complete 
copula descriptions of dependence in ecology will be substantial.

@@

We give a brief introduction to copulas, focusing, for simplicity, on bivariate copulas
and on concepts needed
for the rest of the paper. See @nelsen2006_copula, @joe2014_dependence, and
@MaiScherer2017 for general background, and see also @Anderson2018, part of 
which is a short introduction to 
copulas for ecologists. A bivariate *copula* can be defined as a bivariate distribution
function with both margins uniform on $(0,1)$ (@joe2014_dependence, p. 7). 
This will be the
distribution function of a bivariate random variable with uniform marginals, and 
the terminology *copula* is sometimes also applied to the random variable. 
A foundational theorem of
@sklar1959_theorem (see also @MaiScherer2017, p. 16) 
says that if $F$ is the bivariate distribution function
of a random vector $(X,Y)$, with margins $F_X$ and $F_Y$, then there 
exists a copula $C$
such that for all $(x,y)$ in the Euclidean plane, $F(x,y)=C(F_X(x),F_Y(y))$.
The theorem also says $C$ is unique if $F_X$ and $F_Y$ are continuous.
Thus $C$ couples the bivariate distribution, $F$, with the two marginal distributions. 
Finally, Sklar's theorem says that if $D$ is any bivariate copula and 
$G_A$ and $G_B$ are univariate distribution functions, then 
$D(G_A(x),G_B(y))$ is a distribution function.
The applications of this study fall within the case in which univariate marginal 
distributions have continuous, strictly 
monotonic distribution functions, and this case is simpler, so we 
henceforth make such an assumption. 
Then, "the" copula (now guaranteed unique) 
associated with $(X,Y)$ is the distribution
function of the random variable $(U,V)=(F_X(X),F_Y(Y))$ 
(Appendix \ref{ContStrictMonoton}). And 
$D(G_A(x),G_B(y))$ is the distribution function of the random variable 
$(G_A^{-1}(U),G_B^{-1}(V))$, where $(U,V)$ is the random
vector with distribution function $D$ and
$G_A^{-1}$ and $G_B^{-1}$ are the inverses of $G_A$ and $G_B$
(Appendix \ref{ContStrictMonoton}).

Thus Sklar's theorem implies that
any random vector can be constructed from a copula and marginal 
distributions. And conversely,
any copula and any univariate distribution functions give rise to a random vector. 
Marginals contain no information about the
dependence structure of a random vector, so the copula contains
all such information - it is a complete description of the dependence
between variables. Sklar's theorem makes it possible to construct
bivariate distributions in two separate steps: by specifying the marginal 
distributions and by specifying the dependence structure. The theorem
also make it possible to study and model dependence independently 
from marginal distributions. For instance, in a pedagogical 
example, [@Anderson2018] modelled the dependence between abundance
data for two fish species, red moki and black angelfish, by first 
modelling the marginal distributions and then separately modelling 
the dependence using Gaussian copulas (their section 2).
See the Discussion for a few words on multivariate copulas, which can
be developed in the same way as above [@nelsen2006_copula; @joe2014_dependence;
@MaiScherer2017; @Anderson2018].
<!--***DAN: dont forget to put these few words in!-->

<!--***DAN: I still feel like we have to make this more intuitive for people
at this point. Maybe a figure with two panels showing two copulas on
the left, and two panels shwoing two sets of marginals at the top,
and then four panels showing what you get when each of these combine?
Good idea. Steps: split the current Fig 1 so the new Fig 1 has the first four panels,
but C-D should only have the points. Then make a Fig 2 as above, and edit the
text in the above para to use that figure. 
The rest of the panels of the current Fig 1 will make a new Fig 3 later.-->

<!--***DANr: Add a new section called Background on copulas.
It can have the following kinds of stuff (probably not exhaustive, and probably
not in order, and some of this material is currently below):
1) DONE - The definition, Sklar's theorem
2) What good are copulas - intuitive idea, basically, but also have to be linked to 
the formalities of the definition

What some of the different copula families look like
The explanation of how ranks relate to copulas and estimating the copula parameters
How to carry out fitting (basically, MLE fitting on ranks - Genest et al paper)
What tail dependence actually is
-->

<!--***DAN: In the Discussion, you need to point out not only that we
use only bivariate copulas (and there are multivariate ones), but also
that we use only continuous marginals. Probably an adequate approximation
for our data, but see the other papers I've seen recently which use count data,
and discrete marginals. Approaches such as that will be necessary for some 
datasets-->

@@

<!--***DANr: I moved some text from here and have not smoothed the transtion,
if it needs smoothing-->

### 2019-09-07 SupportingInformation.Rmd
some progress writing a Background section

>>>>

# Proofs of statements related to Skalr's theorem in the case of continuous, strictly monotonic margins \label{ContStrictMonoton}

Let $F$ be the bivariate distribution function
of a random vector $(X,Y)$, with margins $F_X$ and $F_Y$ assumed to be 
continuous and strictly monotonic. We show the copula associated with
$(X,Y)$, which exists and is unique by Sklar's theorem, is the distrbution function
of the random variable $(U,V)=(F_X(X),F_Y(Y))$. Let $C$ denote the distribution
function of $(U,V)$, which has uniform margins. By the uniqueness
statement of Sklar's theorem, it suffices to show that $F(x,y)=C(F_X(x),F_Y(y))$
for all $(x,y)$ in the Euclidean plane. But 
$C(F_X(x),F_Y(y)) = P[U \leq F_X(x), V \leq F_Y(y)]
= P[F_X(X) \leq F_X(x), F_Y(Y) \leq F_Y(y)]
= P[X \leq x, Y \leq y]
= F(x,y),$
where the third equality follows from the strict monotonicity assumption. In fact, 
that equality should hold even with non-strict monotonicity, so a stronger 
statement is possible (see below for more on this topic). The continuity assumption 
for $F_X$ and $F_Y$ is necessary, however: without it the random variables 
$U$ and $V$ are not uniformly distributed, so their joint distribution function 
is not even a copula.

Now let $D$ be any bivariate copula and let 
$G_A$ and $G_B$ be univariate distribution functions, so that 
$D(G_A(x),G_B(y))$ is guaranteed by Sklar's theorem to be a a distribution function.
We here show that if $G_A$ and $G_B$ and continuous and strictly monotonic, then
$D(G_A(x),G_B(y))$ is the distribution function of the random variable 
$(G_A^{-1}(U),G_B^{-1}(V))$, where $(U,V)$ is the random
vector with uniform marginals associated with $D$ and
$G_A^{-1}$ and $G_B^{-1}$ are the inverses of $G_A$ and $G_B$.
The inverses exist because of the strict monotonicity assumption.
To see this, note that the distribution function of 
$(G_A^{-1}(U),G_B^{-1}(V))$ is $P[G_A^{-1}(U) \leq x, G_B^{-1}(V) \leq y]$,
which equals $P[U \leq G_A(x), V \leq G_B(y)]$, by strict monotonicity.
But this is $D(G_A(x),G_B(y))$ because $D$ is the distribution function of 
$(U,V)$, which proves the statement. In fact, a stronger statement assuming
neither continuity nor strict monotonicity is possible, if one uses generalized
inverses for $G_A^{-1}$ and $G_B^{-1}$ [@MaiScherer2017]. We adopt the weaker
statements for simplicity and because our applications are solely under the case
of continuous, strictly monotonic margins. But see @Anderson2018 and 
@Popovic2018 for applications of copulas outside this case.
